A Meta-Analysis of Music Emotion Recognition Studies

### 2025-03-26 manuscript/manuscript.qmd
manuscript: updates to introduction

In the early 2000s, several research groups conducted studies using regression [@friberg_automatic_2002; @liu_automatic_2003] and classification [@lu_automatic_2005; @feng_popular_2003; @mandel_support_2006] techniques to predict emotion in music audio or MIDI. Citing "MIR researchers' growing interest in classifying music by moods" [@downie_music_2008, p. 1], the Music Information Retrieval EXchange (MIREX) introduced Audio Mood Classification (AMC) to their rotation of tasks in 2007. In the first year, nine systems classified mood labels in a common data set, reaching 52.65% accuracy (SD = 11.19%). These events, along with growing interest in the burgeoning field of affective computing, would lead to an explosion of interest in MER research.[^1]

[^1]: The best-performing model to date reached 69.83 % in the 2017 competition [@park2017representation].

Researchers have assessed regression and classification techniques on diverse corpora with features drawn from music (e.g., audio, MIDI, metadata) and participants (e.g., demographic information, survey responses, physiological signals, etc.). In one widely-cited study, Yang (2008) approached MER as a regression task, predicting the valence (i.e., the negative---positive emotional quality) and arousal (i.e., the calm---exciting quality) of 195 Chinese pop songs, achieving 62% accuracy for arousal but only 28% for valence [@yang2008]. This difference in prediction accuracy between dimensions has reappeared in several subsequent studies [e.g., @bai2016dimensional; @coutinho2013psychoacoustic], with some research suggesting this challenge reflects fewer well-established predictors and more individual differences for valence than arousal [@yang2007music; @eerola2011c].

## The semantic gap in MER

The difficulty in predicting valence reflects a broader challenge in information retrieval. Specifically, relations between low-level predictors from music and text and the perceptual phenomena they model remain poorly understood, reaching a ceiling in prediction accuracy [@celma_foafing_2006]. To address this so-called *semantic gap*, researchers have attempted to identify new feature sets with greater relevance to emotion [@panda2020audio; @chowdhury2021perceived], combine low-, mid-, and high-level features using multimodal data [@celma_foafing_2006], or train neural networks to automatically learn features from audio [@zhang2016br;@alvarez2023ri;@agarwal2021an]. Through these approaches, MER researchers attempt to shatter a so-called *glass ceiling* [@downie_music_2008] by establishing central predictors for emotions. To date, however, no study has systematically compared results of these diverse approaches.

>>>>

## Classification and regression approaches

In the early 2000s, several research groups conducted studies using regression [@friberg_automatic_2002; @liu_automatic_2003] and classification [@lu_automatic_2005; @feng_popular_2003; @mandel_support_2006] techniques to predict emotion in music audio or MIDI. Citing "MIR researchers' growing interest in classifying music by moods" [@downie_music_2008, p. 1], the Music Information Retrieval EXchange (MIREX) introduced Audio Mood Classification (AMC) to their rotation of tasks in 2007. In the first year, nine systems classified mood labels in a common data set, reaching 52.65% accuracy (SD = 11.19%). These events, along with growing interest in the burgeoning field of affective computing, would lead to an explosion of interest in MER research.

<!--Paragraph on MER as regression task-->

In the tenth annual AMC task, the highest performing model reached 69.83% accuracy [@park2017representation]. In parallel, other researchers began assessing MER as a regression task. The first study to popularize this approach predicted valence (i.e., the negative---positive emotional quality) and arousal (i.e., the calm---exciting quality) in 195 Chinese pop songs [@yang2008] using audio-extracted features. The regression model predicted arousal with 62% accuracy, but only achieved 28% accuracy for valence. This difference in prediction accuracy between dimensions has reappeared in several subsequent studies [e.g., @bai2016dimensional; @coutinho2013psychoacoustic], with some research suggesting this challenge reflects fewer well-established predictors and more individual differences for valence than arousal [@yang2007music; @eerola2011c].

<!--paragraph on models used across regession & classification tasks-->

Across regression and classification paradigms, a wide range of models have been employed, ranging from linear regression models [@saizclar2022pr; @griffiths2021am; @yang2008] to deep neural networks [@orjesek2022en; @hizlisoy2021mu]. In classification tasks, early work has commonly employed Gaussian mixture models [@lu_automatic_2005; @liu_automatic_2003] and support vector machines [@mandel2007labrosa; @lin2009eeg; @tzanetakis2007marsyas], whereas convolutional, recurrent, and fully-connected neural networks have become increasingly popular in recent years [@grekow2021music; @song2018audio; @coutinho2017sh]. In regression tasks, a wide range of algorithms have been tested, including partial least squares [@gingras2014be; @wang2021ac], support vector machines [@grekow2018au; @hu2017cr; @agarwal2021an], random forests [@beveridge2018po; @xu2021us], and convolutional neural networks [@orjesek2022en].

## Large vs. small data

MER tasks commonly apply regression and classification techniques to diverse datasets with features drawn from both music (e.g., audio, MIDI, metadata) and participants (e.g., demographic information, survey responses, physiological signals, etc.). To facilitate model comparison, several databases have been shared publicly, including MediaEval [@soleymani20131000], DEAM [@aljanaki2017developing], and AMG1608[@chen2015amg1608]. These datasets represent Western pop music, are moderate in terms of the size (containing from 744 to 1802 music excerpts) and have been manually annotated by a relatively large number of participants (either by experts, students, or crowdsourced workers). Many datasets include audio features analyzed with software toolboxes for audio analysis such as OpenSMILE [@] and MIRToolbox [@lartillot2007matlab]---enabling models that predict ratings using tens, or even hundreds, of audio features. Large datasets are ideal for generalizing predictions across diverse musical samples and are commonly employed in machine learning applications focused on maximizing accuracy of a MER system.

Although training MER systems with a wide range of stimuli and features can be useful for generalizable and accurate modeling, small datasets can be more appropriate in explorations that require precise control of stimuli or analyzed features to understand a model's performance. For example,  @griffiths2021am evaluated a MER model's performance when predicting 20 tracks different genres, validating the model on another 40 tracks. @saizclar2022pr evaluated how onset curves (changes in the energy of a note starting from its onset) can be incorporated into a predictive framework using 40 short (8-12s) piano excerpts. Smaller datasets are also useful in studies attempting to explain how stimulus qualities such as amplitude normalization [@gingras2014be] or different performers' interpretations [@battcock2021in] affect ratings of valence and arousal. Although these studies are not always designed with MER applications in mind, their similar framework has enabled direct application [e.g., @chowdhury2021perceived].

## The semantic gap in MER / What are the reported standards? / ceiling

```{=html}
<!--
- Wide range of MER models attempting to outperform previous benchmarks, comparing performance against baseline models
- Introduce ceilings for classification, valence, and arousal, citing Barthet study
- Citing what some researchers attribute limitations to (e.g., lower valence than arousal)
- MIR discussions of "semantic" challenge
- inherent error in all stages (annotation, features, models)
- Due to wide range of approaches, difficult to evaluate how successful attempts to raise glass ceiling have been
- How do modeling techniques, and the number of features used affect performance? Which combinations perform best?

-->
```

Researchers have attempted to identify new feature sets with greater relevance to emotion [@panda2020audio; @chowdhury2021perceived], combine low-, mid-, and high-level features using multimodal data [@celma_foafing_2006], or train neural networks to automatically learn features from audio [@zhang2016br;@alvarez2023ri;@agarwal2021an]. Through these approaches, MER researchers attempt to shatter a so-called *glass ceiling* [@downie_music_2008] by establishing central predictors for emotions. To date, however, no study has systematically compared results of these diverse approaches.
